\documentclass{article}

\usepackage{geometry}
\geometry{a4paper,total={190mm,257mm},left=10mm,top=15mm}

\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{colorlinks=true,urlcolor=blue}
\usepackage{tablefootnote}

\pagestyle{empty}

\begin{document}
	
\begin{center}
	\section*{{\huge Curriculum vitae}}
\end{center}

\hfill

\begin{minipage}[c]{.45\textwidth}
	\subsection*{Personal information}
	\begin{itemize}
		\item Name: \textit{Tomáš Turek}
		\item Date of birth: \textit{24.03.2001}
		\item Origin: \textit{Olomouc, Česká republika}
		\item Mail: \href{mailto:tom.turek@atlas.cz}{\textit{tom.turek@atlas.cz}}
		\item Github: \href{https://github.com/metury}{\textit{metury}}
		\item Drivers license: \textit{B}.
	\end{itemize}
\end{minipage}
\hfill
\begin{minipage}[c]{.5\textwidth}
	\subsection*{Technology and programming languages}
	\begin{itemize}
		\item Advanced knowledge: \textit{C++} and \textit{Qt}.
		\item Average knowledge: \textit{Java}, \textit{Python} and \textit{Rust}.
		\item Basic knowledge: \textit{C\#}, \textit{Haskell} and \textit{Perl}.
	\end{itemize}
	
	\subsection*{Languages}
	\begin{itemize}
		\item Advanced knowledge: \textit{Czech}, \textit{English}.
		\item Basic knowledge: \textit{German}.
	\end{itemize}
\end{minipage}

\subsection*{Previous studies}

\begin{table}[!ht]\centering
	\begin{tabular}{p{20mm} p{50mm} p{80mm} p{20mm}}
		\textit{Years} & \textit{Type} & \textit{School} & \textit{Ended with} \\
		2023-2025\tablefootnote{Expected end of studies.} & Discrete models and algorithms & \href{https://www.mff.cuni.cz/}{Univerzita Karlova, Matematicko-Fyzikální fakulta} & -- \\
		2020-2023 & Computer science & \href{https://www.mff.cuni.cz/}{Univerzita Karlova, Matematicko-Fyzikální fakulta} & Bc. title \\
		2012-2020 & High school & \href{https://www.gytool.cz/}{Gymnázium Olomouc-Hejčín} & Graduation \\
		2007-2012 & Primary school & \href{https://zsstupkova.cz/}{ZŠ Stupkova, Olomouc} & \\
	\end{tabular}
\end{table}

\subsection*{Projects I have worked on}

\begin{itemize}
	\item Application \href{https://rodoc-app.github.io/}{rodoc} for organizing family trees, written in \texttt{C++} with usage of \texttt{Qt}. This was part of my bachelors thesis.
	\item Another smaller project is my solutions for \href{https://adventofcode.com/}{advent of code}, which also has a script written in \texttt{Perl} for automatic projects and page generations -- \href{https://github.com/metury/advent-of-code}{github repository}.
\end{itemize}

\subsection*{Previous work experiences}

\begin{table}[!ht]\centering
	\begin{tabular}{p{20mm} p{150mm}}
		\textit{Years} & \textit{Work} \\
		2019 \& 2020 & Summer manual job in Denmark -- sorting and preparing of packages in a logistic company Prime Cargo.
	\end{tabular}
\end{table}

\subsection*{Personal characteristics}

\begin{itemize}
	\item I'm happy to learn new technologies and practices because nothing stays the same in the world of IT.
	\item Coming up with solutions to problems that occur on a daily basis, whether at work or in personal life.
	\item I consider myself a hardworking person who is willing and happy to help others.
	\item Besides, I like to be in the company of other people. And I'm also a very punctual.
	\item Since no one is perfect, I also have some disadvantages, such as occasional stubbornness.
\end{itemize}

\subsection*{Free time hobbies}

\begin{itemize}
	\item I played floorball competitively for several years and I still play it sometimes when the opportunity arises.
	\item Also playing the flute. I learned it while attending the \href{https://www.zus-zerotin.cz/}{Žerotín Music School}, where, apart from playing solo, I also played in smaller groups and also in bigger groups, with which we competed, among other things, in foreign festivals.
\end{itemize}

\end{document}
