\chapter{Balls \& Bins}

This model is if we have $m$ balls and $n$ bins and for each ball we put it independently at random to one bin, where each bin has the same probability.

One well known problem is \textit{Birthday paradox} where we have $k$ people as balls and $365$ days as bins. Then we are asking what is the probability that one bin has at least 2 balls.

$$
Pr[\text{at least 2 balls in one bin}] = 1 - Pr[\text{max 1 ball in one bin}] =
$$

$$
= 1 - \prod_{i=1}^{m-1} \frac{n-i}{n} \approx 1 - \prod_{i=1}^{m-1} e^{\frac{-i}{n}} = 1 - e^{\frac{-m(m-1)}{2n}}
$$

We also consider other properties, such as the expected number of empty bins:

$$
Pr[\text{bin } i \text{ is empty}] = \left(1 - \frac{1}{n}\right)^{m} \approx e^{\frac{-m}{n}}
$$

$$
\mathbb{E}[ \text{\# of empty bins}] = n \left(1- \frac{1}{n}\right)^{m} \approx n e^{\frac{-m}{n}}
$$

\begin{thm}[Max Load Theorem]
	If $m=n$ and are big enough and $M = 3 \frac{\ln(n)}{\ln\left(\ln\left(n\right)\right)}$ then
	
	$$
	Pr[\text{max \# of balls in a bin} > M] < \frac{1}{n}.
	$$
\end{thm}

\begin{proof}
	% TODO: I don't think this is right. The proba of the Bin distribution should be smaller not greater than >= M balls
	
	$$
	Pr[\text{bin } \#1 \text{ has} \geq M \text{ balls} ]  \leq Pr\left[Bin\left(n, \frac{1}{n}\right) = M\right] < \frac{1}{M!} < \left(\frac{e}{M}\right)^M
	% The binomial distribution is wrong in this case
	$$
	
	$$
	Pr[\text{any bin has} \geq M \text{ balls} ] \leq
	$$
	$$
	\leq Pr[\text{bin } \#1 \text{ has} \geq M \text{ balls} ] + \dots + Pr[\text{bin } \#n \text{ has} \geq M \text{ balls} ] \leq n \left(\frac{e}{M}\right)^M
	$$
	
	Now we will show that this expression is smaller than $\frac{1}{n}$. In order to do that, we wil take the logarithm of both sides and we get:
	
	$$
	2\ln(n) + M\left(1-\ln\left(M\right)\right) < 0
	$$
	
	We will then substitute $M$ and show that the inequality holds.
\end{proof}


M balls and bins have multiple applications. We will use it for hashing and sorting.

\section{Bucket Sort Application}

We want to sort $n = 2^k$ numbers from range $[0, 2^l-1]$ where $l > k$. The numbers are uniformly random in this range.

\subsection{Algorithm}

\begin{enumerate}
	\item Put input $x$ to a bucket $b(x)$ where $b(x)$ is a hash function of $x$ and bucket is a list.
	\item Sort each bucket (list) by a bubble sort in quadratic time.
	\item Merge the buckets.
\end{enumerate}

\subsubsection{Time Analysis}

Parts 1 and 3 are linear in $n$. For part 2, we will consider $X_i = \# \text{of inputs in the ith bucket} \sim Bin(n, \frac{1}{n})$. Then $\mathbb{E} \text{ time} = \mathbb{E} \sum (c_iX_i^2)$. Finally, we will use the definition of variance to show that the expected time is $ < 2cn$. Hence the whole algorithm has linear expected time.

\section{Hash Collisions Application}

We want to store $n$ strings and search fast. Using the max load theorem, we will show that max running time with a big enough $n$ is $< 3 \frac{\ln(n)}{\ln(\ln(n))}$.

\begin{thm}
	Distribution of $X^{(m)}_{1}, \dots, X^{(m)}_n$, where $X_i^{(m)} $ represents the number of balls in bin $i$, is the same as $Y^{(m)}_1, \dots Y^{(m),}_n \ iid$, where $Y^{(m)}_i \sim Pois\left(\frac{m}{n}\right)$ and $\sum Y_i^{(m)} = k$
\end{thm}

\begin{proof}
	It is based on the fact that $X^{(m)}_1 \sim Bin(m, \frac{1}{n}) \approx Pois(\frac{m}{n})$ and then we show that $Pr[X^{(m)}_1 = k_1, \dots X^{(m)}_n = k_n] = P_x = P_Y = Pr[Y^{(m)}_1 = k_1 \dots \vert \sum Y_i = k ]$
\end{proof}

\begin{thm}[Max Load Theorem 2]
	If $m=n$ and are big enough and $M = \frac{\ln(n)}{\ln\left(\ln\left(n\right)\right)}$ then
	
	$$
	Pr[\text{max \# of balls in a bin} < M] \leq \frac{1}{n}.
	$$
\end{thm}
